---
title: "R Notebook"
output: html_notebook
---

Intro to stats: https://docs.google.com/document/d/1_7XSn_qEFHdGaQnfEauRkOkhtvHNS4eP9mj-ATK4GkI/edit

Welcome to R! R is essentially a programming language for statistics - everything is built for data analysis! In this workshop, we'll be going through some of the basics of R. 
Unfortunately, we won't be able to go through absolutely every facet of R, but if you have any questions, feel free to ask us at the end of the workshop!
In order to use R, you first need to install the datasets you're using. Here's how you do that:
```{r}
# Installing packages
install.packages("covid19.analytics")
install.packages("gapminder")

#loading Packages
library(covid19.analytics)
library(gapminder)

# final initialization
data("gapminder") #this may not be necesscary, depending on your data, but check the docs for your dataset!
```

```{r}
# Summary gives you a breif overview of the data
summary(gapminder)
```

```{r}
#the syntax may depend on the particular library you're using
report.summary()
```

```{r}
#use the $ to select a specific value from your data
# your data may have a different structure though - make sure you read the documentation!
mean(gapminder$lifeExp)
```

```{r}
# use <- to assign particular data to variables

x <- mean(gapminder$lifeExp)
all.data <- covid19.data('ts-ALL')
```
```{r}
#if you're feeling efficient, you can use attach to make one dataset your default
attach(gapminder)

median(pop)
#now we don't need to keep calling gapminder!
```
```{r}
#r has a few built in functions
hist(lifeExp) #makes a histogram
boxplot(lifeExp) #boxplot
```
```{r}
#you can also do log plots if your data are skewed

hist(pop)

#versus

hist(log(pop))
```
```{r}
#use the ~ to create dependent and independent variables
boxplot(lifeExp ~ continent) #continent is the independent vairable
```
```{r}
# you can also create scatterplots
plot(lifeExp ~ log(gdpPercap)) #gdp is independent here
```
These are just the basic functions of R - in order to take full advantage of it, we have to install a few new libraries
Two good ones are dyplr and ggplot, but your dataset may come with built in features such as visualization, so be sure to check for them!
For example...
```{r}
TS.data <- covid19.data("ts-ALL")

totals.plt(TS.data)
```
Here's how to use dplyr
```{r}
# Installing
install.packages("dplyr")
library(dplyr)
```
```{r}
gapminder %>% # the weird percent thingy is called the pipe operator - it basically operates as 'and then'
  select(country, lifeExp) %>% #selecting the country and its life expectency 
  filter(country == "Canada" | country == "South Africa") %>% # the | character works like 'or'
  group_by(country) %>%
  summarise(Average_life = mean(lifeExp)) #this creates our table - the 'dependent' column is assigned to that country's average life expectency
```
Data tables are nice, but you also need to do statistical analysis on your data. Here's how to do a T-test for a difference of means with your data.
So, there's a substantial difference between the life expectancy of Canadians and South Africans. Is this just chance, or is there something more here?

```{r}
data_frame_1 <- gapminder %>% 
  select(country, lifeExp) %>% 
  filter(country == "Canada" | country == "South Africa")
# Taking our data from before and assigning it to our dataframe
```
```{r}
t.test(data = data_frame_1, lifeExp ~ country) #we take our previous dataframe and tell the computer to do a t-test comparing the life expectancies of Canada and South Africa
```
So, based on the P-value we got, we can say that this is statistically significant. This function also gives you a handy confidence interval of what the difference could be!

If you want to create plots, ggplot is the way to do it! Here's how its used.
```{r}
#installing it
install.packages("ggplot2")
library(ggplot2)
```
```{r}
#now let's use it!
gapminder  %>%
  filter(gdpPercap < 50000) %>% #here, we take our gapminder data and filter it by countries with a GDP per capita of less than 50000
  ggplot(aes(x=log(gdpPercap), y=lifeExp, col=continent, size=pop))+geom_point()
#There's a lot to unpack here - first, we tell it to put Gdp on the x axis and life expectancy on the y axis, and then colour by continent, and finally make the bubbles proportional to population!

```
There's a lot more you can do with ggplot, but I highly encourage you to explore on your own! There's a lot of potential for great visualizations.
Another useful model is a linear regression, which can be used to essentially create trendlines. Here's how we do that.
```{r}
lm(lifeExp ~ gdpPercap) #This model asks how much GDP impacts life expectancy
```
Cool! That shows that there is a correlation between the two in our data. Here's some more advanced linear modelling!
```{r}
summary(lm(lifeExp ~ gdpPercap))
#this gives more in depth stats, such as p-values and residuals
```
```{r}
lm(lifeExp ~ gdpPercap+pop)
#we can even do some multivariable models!
```

I hope this is enough to get you started on your journey in R! If you have any questions, please ask us! After all, we're here to help!
